PAPER DRAFT

### 2021-11-04 Papers/Polymethylation_scores/06_discussion.Rmd
more revisions post Kelly 1st round comments

In the longitudinal Fragile Families and Child Wellbeing birth cohort, we observed prenatal maternal smoking was associated with DNA methylation in children's saliva samples from ages 9 and 15. Prenatal maternal smoking was associated with polymethylation scores across strata of child age and ancestry. Global methylation and epigenetic clocks were not associated with maternal smoking exposure demonstrating a specific and persistent signal of prenatal smoke exposure in DNA methylation. Polymethylation scores had reasonable accuracy for classifying prenatal maternal smoking. This accuracy was more consistent across time than when using individual *a priori* CpG sites.

Our findings are consistent with the previous literature on associations between prenatal maternal smoking and DNA methylation. We replicated top hits from previous epigenome wide association studies, including a large meta-analysis of DNA methylation in newborn blood samples [@Joubert:2016ue]. There are no previous studies of prenatal maternal smoking and saliva DNA methylation. However, previous work has that the majority of CpG sites were similarly methylated in blood and saliva (reviewed in @Langie2017). DNA methylation at AHRR:cg05575291 can accurately classify own--smoking behavior in both blood (area under the curve 0.995) and saliva (area under the curve 0.971) [@Dawes2019; @Philibert2020].

We advance the literature on prenatal maternal smoking and DNA methylation by evaluating the persistence of the association as children age and its portability across tissue and ancestry. Polymethylation scores built using coefficients from meta-analysis of cord blood DNA methylation from primarily European-ancestry newborns were still associated with prenatal maternal smoking in our independent saliva samples from a diverse cohort at ages 9 and 15 [@Joubert:2016ue]. The portability of other risk scores, such as polygenic risk scores, across ancestries is a complex research area [@Martin2019; ] and evaluating the portability of epigenetic summary measures has been identified as a key area for evaluation [@Hüls2020]. In this case, the polymethylation score for prenatal maternal smoking appears to be portable across ancestry groups. Effect estimates were consistent across ancestries at age 9, although Hispanic and European genetic ancestry samples had lower effect estimates at age 15. This could reflect higher unreported smoking initiation in European and Hispanic ancestry teens in the United States than African ancestry teens. Genetic ancestry correlates with race, and White and Latino teens have much higher rates of teen smoking and earlier ages at initiation than Black children [@El-Toukhy2016]. Though we excluded children who reported own-smoking from our analytic sample, under-reporting of own-smoking could influence DNA methylation at ages 9 and 15, creating outcome misclassification. This could also explain the observed attenuation the effect on AHRR:cg05575921 methylation by age 15 in our sample, as DNA methylation at this probe is known to vary by personal cigarette smoking[@Dawes2019; @Philibert2020].

Our findings are also consistent with previous work on DNA methylation as a biomarker for prenatal maternal smoke exposure. Prenatal maternal smoking classification using saliva DNA methylation in our sample of 9 and 15 year olds () performed comparably to classification using peripheral blood from older adults () [@Richmond:2018tu]. However, previous biomarkers of prenatal maternal smoking using DNA methylation of cord blood from newborns () and peripheral blood from 3-5 year old children () more accurately classified prenatal maternal smoke exposure [@Reese:2017ts; @Ladd-Acosta:2016aa]. Differences in DNA methylation patterns across tissues and over time may influence the performance of DNA methylation biomarkers. The similarity in performance between saliva DNA methylation in our study and peripheral blood DNA methylation in older adults is encouraging for the use of saliva DNA methylation as a readily accessible clinical sample.

Our analysis contributes to the development of an accurate methylation biomarker for prenatal maternal smoking by evaluating the impact of specific methodological choices on accuracy of classification. DNA methylation biomarkers may be especially susceptible to confounding by subject age at methylation measurement, tissue of measurement, and cell-type proportion [@Hüls2020]. Accuracy of polymethylation score classification was similar when using coefficients from cord blood vs peripheral blood samples from older children. Accuracy was not improved by using two time-points of methylation measurement. Accuracy was improved when polymethylation scores used coefficients which incorporated cell-type control. Our analysis therefore suggests that cell-type control when both generating and applying coefficients for polymethylation scores may positively influence accuracy, although more work comparing coefficients from different populations is still needed.

Our analysis suggests that polymethylation scores may be more accurate than using single CpG sites as biomarkers. AHRR:cg05575921 is a consistent marker of prenatal maternal smoke exposure in meta-analyses of newborn cord blood[@Joubert:2016ue] and a very accurate DNA methylation marker of own smoking [@Philibert2020]. However, salivary AHRR:cg05575921 was not a persistent marker of prenatal maternal smoking in an analysis of middle-aged adult women [@Richmond:2018tu]. In our sample, salivary AHRR:CpG05575921 methylation was less accurate at predicting prenatal maternal smoke exposure than polymethylation scores. The accuracy of AHRR:CpG05575921 and other single CpG site biomarkers may be influenced by time-since-exposure and new environmental exposures. Incorporating information across multiple sites of DNA methylation may yield a biomarker more robust to these influences.

>>>>

In the longitudinal Fragile Families and Child Wellbeing birth cohort, we observed that prenatal maternal smoking was associated with DNA methylation in children's saliva samples from ages 9 and 15. Prenatal maternal smoking was associated with polymethylation scores across strata of child age and ancestry. Global methylation and epigenetic clocks were not associated with maternal smoking exposure demonstrating a specific and persistent signal of prenatal smoke exposure in DNA methylation. Polymethylation scores had reasonable accuracy for classifying prenatal maternal smoking. This accuracy was more consistent across time than when using individual *a priori* CpG sites.

Our findings are consistent with the previous literature on associations between prenatal maternal smoking and DNA methylation. We replicated top hits from previous epigenome wide association studies of DNA methylation in cord and peripheral blood[@Joubert:2016ue; @wiklund2019]. There are no previous studies of prenatal maternal smoking and saliva DNA methylation. However, previous work has shown that the majority of CpG sites are similarly methylated in blood and saliva (reviewed in @Langie2017).

We advance the literature on prenatal maternal smoking and DNA methylation by evaluating the persistence of the association as children age and its portability across tissue and ancestry. Polymethylation scores built using coefficients from meta-analysis of cord blood DNA methylation from primarily European-ancestry newborns were still associated with prenatal maternal smoking in our independent saliva samples from a diverse cohort at ages 9 and 15 [@Joubert:2016ue]. The portability of other risk scores, such as polygenic risk scores, across ancestries is a complex research area [@Martin2019] and evaluating the portability of epigenetic summary measures has been identified as a key area for evaluation [@Hüls2020]. In this case, the polymethylation score for prenatal maternal smoking appears to be portable across ancestry groups. Effect estimates were consistent across ancestries at age 9, although Hispanic and European genetic ancestry samples had lower effect estimates at age 15. This could reflect higher unreported smoking initiation in European and Hispanic ancestry teens in the United States than African ancestry teens. Genetic ancestry correlates with race, and White and Latino teens have much higher rates of teen smoking and earlier ages at initiation than Black children [@El-Toukhy2016]. Though we excluded children who reported own-smoking from our analytic sample, under-reporting of own-smoking could influence DNA methylation at ages 9 and 15, creating outcome misclassification. This could also explain the observed attenuation the effect on AHRR:cg05575921 methylation by age 15 in our sample, as DNA methylation at this probe is known to vary by personal cigarette smoking[@Dawes2019; @Philibert2020].

Our findings are also consistent with previous work on DNA methylation as a biomarker for prenatal maternal smoke exposure. Prenatal maternal smoking classification using saliva DNA methylation in our sample of 9 and 15 year olds performed comparably to classification using peripheral blood from older adults (AUC 0.72) [@Richmond:2018tu]. However, previous biomarkers of prenatal maternal smoking using DNA methylation of cord blood from newborns (AUCs ranging from 0.82-0.97) and peripheral blood from 3-5 year old children (AUC=0.87) more accurately classified prenatal maternal smoke exposure [@Reese:2017ts; @Ladd-Acosta:2016aa]. Differences in DNA methylation patterns across tissues and over time may influence the performance of DNA methylation biomarkers. The similarity in performance between saliva DNA methylation in our study and peripheral blood DNA methylation in older adults is encouraging for the use of saliva DNA methylation as a readily accessible clinical sample.

Our analysis contributes to the development of an accurate methylation biomarker for prenatal maternal smoking by evaluating the impact of specific methodological choices on accuracy of classification. DNA methylation biomarkers may be especially susceptible to confounding by subject age at methylation measurement and cell-type proportion [@Hüls2020]. Accuracy of polymethylation score classification was similar when using coefficients from cord blood vs peripheral blood samples from older children. Accuracy was not improved by using two time-points of methylation measurement. Accuracy was improved when polymethylation scores used coefficients which incorporated cell-type control. Our analysis therefore suggests that cell-type control when both generating and applying coefficients for polymethylation scores may positively influence accuracy, although more work comparing coefficients from different populations is still needed.

Our analysis suggests that polymethylation scores may be more accurate than using single CpG sites as biomarkers. AHRR:cg05575921 is a consistent marker of prenatal maternal smoke exposure in meta-analyses of newborn cord blood[@Joubert:2016ue]. DNA methylation at AHRR:cg05575291 can accurately classify own--smoking behavior in both blood (area under the curve 0.995) and saliva (area under the curve 0.971) [@Dawes2019; @Philibert2020]. However, salivary AHRR:cg05575921 was not a persistent marker of prenatal maternal smoking in an analysis of middle-aged adult women [@Richmond:2018tu]. In our sample, salivary AHRR:CpG05575921 methylation was less accurate at predicting prenatal maternal smoke exposure than polymethylation scores. The accuracy of AHRR:CpG05575921 and other single CpG site biomarkers may be influenced by time-since-exposure and new environmental exposures. Incorporating information across multiple sites of DNA methylation may yield a biomarker more robust to these influences.

### 2021-11-04 Papers/Polymethylation_scores/08_methods.Rmd
more revisions post Kelly 1st round comments

The Fragile Families and Child Wellbeing Study is a birth cohort of nearly 5,000 children born in 20 cities in the United States between 1998 and 2000 [@Reichman:2001wv]. Participants were selected at delivery using a three-stage stratified random sample design which oversampled unmarried mothers by a ratio of 3:1 [@Reichman:2001wv]. Participants were excluded on the following criteria: those with parents who planned to place the child for adoption, those where the father was deceased, those who did not speak English or Spanish well enough to be interviewed, births where the mothers or babies were too ill to complete the interview, and those where the baby died before the interview could take place. Children were followed longitudinally with assessments at ages 1, 3, 5, 9 and 15; additional follow up is ongoing. Assessments included medical record extraction, biosample collection, in-home assessments, and surveys of the mother, father, primary caregiver, teacher and child. At ages nine and fifteen a saliva sample was taken from the child [@Reichman:2001wv]. A subsample of the Fragile Families cohort was selected for saliva DNA and DNA methylation processing. <!--#This subsampling was .... PUT SUBSAMPLE technique here -->

@@

Child covariates included child sex, child report of personal cigarette smoking, and child genetic ancestry. Mothers reported sex (male/female) of their child at baseline. At ages 9 and 15, the child responded to interview questions about ever smoking a cigarette.

@@

We created polymethylation scores for prenatal maternal smoke exposure. From an independent meta-analysis of prenatal smoke exposure and newborn DNA methylation, we extracted the regression coefficients of 6,074 CpG sites associated with prenatal maternal smoking at a false discovery ratio corrected $P$ value \<0.05 [@Joubert:2016ue]. We mean-centered the DNA methylation beta values in our study, weighted them by the independent refression coefficience and took the sum. We calculated polymethylation scores using regression coefficients from 4 different regressions (sustained smoking exposure and DNA methylation in newborn cordblood with cell-type control, sustained smoking exposure and DNA methylation in newborn cordblood without cell-type control, sustained smoking exposure and DNA methylation in peripheral blood from older children, any smoking exposure and DNA methylation in newborn cordblood) [@Joubert:2016ue]. Global DNA methylation was calculated for each sample as the mean methylation value of each sample across the cleaned probe set. Pediatric epigenetic age was calculated for each sample using the coefficients and methods provided by the creators (see [<https://github.com/kobor-lab/Public-Scripts>](https://github.com/kobor-lab/Public-Scripts){.uri}) [@mcewen2019]. The GRIM age clock, including the smoking pack-years subscale, was calculated as previously described [@lu2019].

Single *a priori* CpG sites were selected from the previous large meta-analysis of prenatal maternal smoking and DNA methylation in children's cord and peripheral blood samples [@Joubert:2016ue].

@@

From the 1692 samples from 884 unique individuals with quality controlled DNA methylation data, we further excluded any samples missing data on: maternal prenatal smoking (0 samples), alcohol (4 samples) or other drug use (6 samples), maternal income to poverty ratio (0 samples), maternal postnatal smoking data ( samples). We also excluded samples with missing child sex (0 samples), child age (0 samples) or genetic data ( samples). Finally, if a child reported ever smoking a cigarette at age 9, we excluded their samples. Age 15 samples from children who reported ever smoking a cigarette at age 15 were also excluded.

@@

@@

>>>>

The Fragile Families and Child Wellbeing Study is a birth cohort of nearly 5,000 children born in 20 cities in the United States between 1998 and 2000 [@Reichman:2001wv]. Participants were selected at delivery using a three-stage stratified random sample design which oversampled unmarried mothers by a ratio of 3:1 [@Reichman:2001wv]. Participants were excluded on the following criteria: those with parents who planned to place the child for adoption, those where the father was deceased, those who did not speak English or Spanish well enough to be interviewed, births where the mothers or babies were too ill to complete the interview, and those where the baby died before the interview could take place. Children were followed longitudinally with assessments at ages 1, 3, 5, 9 and 15; additional follow up is ongoing. Assessments included medical record extraction, biosample collection, in-home assessments, and surveys of the mother, father, primary caregiver, teacher and child. At ages nine and fifteen a saliva sample was taken from the child [@Reichman:2001wv]. A subsample of the Fragile Families cohort was selected for saliva DNA and DNA methylation processing. To be eligible for DNA methylation children had to 1) have participated and given saliva at age 9 and 2) participated in the SAND {ACRONYM HELP HERE} study, which oversampled from Detroit, Toledo and Chicago. A random sample of these eligible participants was then selected. <!--#This subsampling was .... PUT SUBSAMPLE technique here -->

@@

Child covariates included child sex, child report of personal cigarette smoking, and child genetic ancestry. Mothers reported sex (male/female) of their child at baseline. At ages 9 children were asked if they had ever smoked a cigarette or used tobacco (yes/no) and at age 15 they were asked if they had ever smoked an entire cigarette (yes/no).

@@

We created polymethylation scores for prenatal maternal smoke exposure. From an independent meta-analysis of prenatal smoke exposure and newborn DNA methylation, we extracted the regression coefficients of 6,074 CpG sites associated with prenatal maternal smoking at a false discovery ratio corrected $P$ value \<0.05 [@Joubert:2016ue]. We mean-centered the DNA methylation beta values in our study, weighted them by the independent regression coefficients and took the sum. We calculated polymethylation scores using regression coefficients from 4 different regressions. For our main analysis we used coefficients from a regression of sustained smoking exposure and DNA methylation in newborn cordblood with cell-type control. As sensitivity analyses, we used re sustained smoking exposure and DNA methylation in newborn cordblood without cell-type control, sustained smoking exposure and DNA methylation in peripheral blood from older children, any smoking exposure and DNA methylation in newborn cordblood) [@Joubert:2016ue]. Global DNA methylation was calculated for each sample as the mean methylation value of each sample across the cleaned probe set. Pediatric epigenetic age was calculated for each sample using the coefficients and methods provided by the creators (see [\<https://github.com/kobor-lab/Public-Scripts\>](https://github.com/kobor-lab/Public-Scripts){.uri}) [@mcewen2019]. The GRIM age clock, including the smoking pack-years subscale, was calculated as previously described [@lu2019].

Single *a priori* CpG sites were selected from previous large meta-analyses of prenatal maternal smoking and DNA methylation in children's cord and peripheral blood samples [@Joubert:2016ue; @wiklund2019].

@@

From the 1692 samples from 884 unique individuals with quality controlled DNA methylation data, we further excluded any samples missing data on: maternal prenatal smoking (0 samples), alcohol (4 samples) or other drug use (6 samples), maternal income to poverty ratio (0 samples), maternal postnatal smoking data ( samples). We also excluded samples with missing child sex (0 samples), child age (0 samples) or genetic data ( samples). Finally, if a child reported ever smoking a cigarette or using tobacco age 9, we excluded all of their available samples. Age 15 samples from children who reported ever smoking a cigarette at age 15 were also excluded. Children who were missing a response to the question at age 9 but answered that they had never smoked a whole cigarette at age 15 were kept in the sample.

@@

#### Receiver operator curve analysis

To evaluate the accuracy of the DNA methylation summary measures as biomarkers of prenatal maternal smoking, we used a receiver operating curve. First, we regressed exposure to prenatal maternal smoking (outcome) againts the DNA methylation summary measures in individual logistic regressions, while adjusting for the base model variables listed above. Next, we calculated receiver operating curves (ROC) and area under the curves (AUCs) using the function roc from the R library pROC version 1.18.0. We compared ROC curves and AUC using the delong method and the function roc.test from the R library pROC version 1.18.0.

@@

In addition to the cross-sectional models within visit-strata, we performed a linear mixed effect model with a random intercept for child.

### 2021-11-04 Papers/Polymethylation_scores/98_declarations.Rmd
added declarations
more revisions post Kelly 1st round comments

## Ethics approval and consent to participate

## Consent for publication

@@

>>>>

### Ethics approval and consent to participate

Participants provided written informed consent for the study. The data used in this manuscript were prepared by the Fragile Families and Childhood Wellbeing Study administrators following approval of the manuscript proposal. These secondary data analyses were approved by the University of Michigan Institutional Review Board (IRB, HUM00129826)

### Consent for publication

@@

### Availability of data and materials

Many of the variables used in this analysis are publicly available in the Fragile Families dataset. Some of the data used in this analysis, including genetic and epigenetic data, is restricted use but available to researchers upon reasonable application.

### Competing interests

The authors have no competing interests to declare. check here

### Funding

This research was made possible through several grants (add). FB was supported by the National Institutes of Health National Institute of Dental and Craniofacial Research (F31DE029992).

### Authors' contributions

FB performed the analysis with contributions from JD, JF, and EW. FB and KB wrote the paper with all authors contributing to revisions. DN, CM, KM, EW and FB all contributed to the design and conception of the study.

### Acknowledgements

### 2021-11-04 Papers/Polymethylation_scores/index_paper.Rmd
more revisions post Kelly 1st round comments

---
title: "Polymethylation scores paper"
author: "F blostein"
date: "9/13/2021"
output:
  word_document: default
  html_document:
    df_print: paged
bibliography: Methylation_SmokingBib.bib
csl: springer-vancouver.csl
always_allow_html: true
---

@@

>>>>

---
title: "PAPER DRAFT"
author: "------------"
date: "`r Sys.Date()`"
output:
  word_document: default
  html_document:
    df_print: paged
bibliography: Methylation_SmokingBib.bib
csl: springer-vancouver.csl
always_allow_html: true
---

@@

```{r child = here::here("Papers", "Polymethylation_scores", "99_FigureCaptions.Rmd"), eval=results_run}
```
